Fractional Types

### 2013-02-04 frac.tex
small change to sec3

We review our language {{langRev}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{langRev}}. Other versions
  include the empty type, recursive types, and trace opeators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions; rather,
the terms are isomorphism witnesses.  In other words, the terms of {{Pi}} are
proofs that certain ``shapes of values'' are isomorphic.  And, in classical
Curry-Howard fashion, our operational semantics shows how these proofs can be
directly interpreted as actions on ordinary values which effect this shape
transformation. Of course, ``shapes of values'' are very familiar already:
they are usually called \emph{types}.  But usually one designs a type system
as a method of classifying terms, with the eventual purpose to show that
certain properties of well-typed terms hold, such as safety.  Our approach is
different: we start from a type system, and then present a term language
which naturally inhabits these types, along with an appropriate operational
semantics.

@@

Types include the unit type {{1}}, sum types {{b1+b2}}, and products types
{{b1*b2}}.  Values includes {{()}} which is the only value of type {{1}},
{{left v}} and {{right v}} which inject {{v}} into a sum type, and
{{(v1,v2)}} which builds a value of product type. But these values should be
regarded as largely ancilliary.  In particular we do not treat the above
values as first-class citizens.  They only occur when we want to observe the
effect of an isomorphism.

@@

For example {{ swap* }} induces a function (akin to):
%subcode{code}
%! language = Haskell
% swapProd (a,b) = (b,a)

@@

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will not present the
semantics because we present a more general version in the next section. We
state without proof that the evaluation of well-typed combinators always
terminates and that {{langRev}} is logically reversible, i.e., that for all
combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have
the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
evaluation of {{c v2}} produces {{v1}}.

@@

We extend {{langRevT}} with fractional types, fractional values, and two
combinators {{eta*_b}} and {{eps*_b}}.  We provide the full syntactic and
semantic definitions for the language below.

\begin{definition}[Syntax of {{langRevT}}]
\label{def:langRevT}
The syntax of types, values, and combinators, is:
%subcode{bnf} include main
% natural numbers, n 
% value types, b ::= 1 | b+b | b*b | 1/b | bool 
% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
%
% combinator types, t ::= b <-> b
% isomorphisms, iso ::= swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist | factor 
%                   &|& foldB | unfoldB 
%                   &|& eta*_b | eps*_b
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
values of type~{{1/b}}. Since there are no values of type {{0}}, this
means that {{1/0}} is also an empty type. Generally, we think of a
value like {{1/false}} as a \emph{first-class constraint} that can
only be satisfied if it is matched with an actual value {{false}}.

>>>>

We review our language {{langRev}} providing the necessary background and
context for our higher-order extension.\footnote{The presentation in this
  section focuses on the simplest version of {{langRev}}. Other versions
  include the empty type, recursive types, and trace operators but these
  extensions are orthogonal to the higher-order extension emphasized in this
  paper.} The terms of {{Pi}} are not classical values and functions; rather,
the terms are isomorphism witnesses.  In other words, the terms of {{Pi}} are
proofs that certain ``shapes of values'' are isomorphic.  And, in classical
Curry-Howard fashion, our operational semantics shows how these proofs can be
directly interpreted as actions on ordinary values which effect this shape
transformation. Of course, ``shapes of values'' are very familiar already:
they are usually called \emph{types}.  But usually one designs a type system
as a method of classifying terms, with the eventual purpose to show that
certain properties of well-typed terms hold, such as safety.  Our approach is
different: we start from a type system, and then present a term language
which naturally inhabits these types, along with an appropriate operational
semantics.

@@

Types include the unit type {{1}}, sum types {{b1+b2}}, and products types
{{b1*b2}}.  Values includes {{()}} which is the only value of type {{1}},
{{left v}} and {{right v}} which inject {{v}} into a sum type, and
{{(v1,v2)}} which builds a value of product type. But these values should be
regarded as largely ancillary.  In particular we do not treat the above
values as first-class citizens.  They only occur when we want to observe the
effect of an isomorphism.

@@

@@

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other. We will not present the
semantics until the next section in the context of the extended language.
For now, we state without proof that the evaluation of well-typed combinators
always terminates and that {{langRev}} is logically reversible, i.e., that
for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}}
we have the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
evaluation of {{c v2}} produces {{v1}}.

@@

Syntactically, the extension with fractional types is simple.

\begin{definition}[Syntax of {{langRevTF}}]
\label{def:langRevT}
The syntax of types, values, and combinators, is:
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b | 1/b 
% values, v ::= () | left v | right v | (v,v) | 1/v 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist | factor 
%                   &|& eta*_b | eps*_b
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. We think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of nothing --- an equal amount of negative and
positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
consumes an equal amount of negative and positive information producing 
no residual information.

give some intuition; example with eta/epsilon

motivate the need for relations

present formal semantics
